---
title: "Getting Started with elicitcausal"
author: "Robert Kubinec"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    highlight-style: github
vignette: >
  %\VignetteIndexEntry{Getting Started with elicitcausal}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>"
)
```

## Overview

**elicitcausal** turns a causal DAG drawn in
[dagitty](https://www.dagitty.net) into a fully specified Bayesian network
by asking you — or your subject-matter experts — for conditional
probabilities one question at a time.

The workflow is:

1. Specify the causal structure as a `dagitty` object.
2. Call `elicit_dag_priors()`. The function walks each node in topological
   order and asks for $P(\text{node} = 1 \mid \text{parent combination})$
   for every possible combination of binary parent values.
3. Receive the elicited conditional probability tables (CPTs), the full
   joint distribution over all $2^n$ states, and Shannon entropy.
4. Query the resulting model with `get_marginal()`, `get_conditional()`, and
   `get_mutual_info()`.

All nodes are assumed **binary** (0 = absent/false, 1 = present/true).

---

## Installation

```{r install, eval=FALSE}
# Install from GitHub (once the package is available)
remotes::install_github("rmk118/elicitcausal")
```

```{r load}
library(elicitcausal)
library(dagitty)
```

---

## A minimal example: X → Y

The simplest possible causal DAG has one cause $X$ and one effect $Y$.

```{r dag-xy}
dag_xy <- dagitty("dag { X -> Y }")
```

In an interactive session you would call `elicit_dag_priors(dag_xy)` and
type probabilities at the prompts. Here we supply them via the `.responses`
argument so the vignette runs without user input.

`X` is a root node, so we need only its **prior** probability. `Y` has `X`
as its single parent, so we need two conditional probabilities: one for
each value that `X` can take.

```{r simple-run}
result_xy <- elicit_dag_priors(
  dag_xy,
  target = "Y",
  .responses = list(
    X = list(0.30),         # P(X = 1) = 0.30
    Y = list(0.10, 0.75)    # P(Y=1|X=0) = 0.10,  P(Y=1|X=1) = 0.75
  )
)
```

The printed output shows:

- **Conditional probability tables** — one per node.
- **Joint distribution summary** — Shannon entropy of the $2^n$-state
  distribution and the top five most probable joint states.
- **Target marginal** — $P(Y = 0)$ and $P(Y = 1)$ after marginalising over
  $X$, plus the marginal entropy $H(Y)$.

---

## Input modes

### Probability mode (default)

In probability mode you type decimal values directly. Any number in $[0,
1]$ is accepted. This is the most precise option and works well when
experts are comfortable with numerical probabilities.

```{r prob-mode, eval=FALSE}
# Interactive usage — not run here
result <- elicit_dag_priors(dag_xy, mode = "probability")
```

```
P(X = 1) = 0.3
P(Y = 1 | X = 0) = 0.10
P(Y = 1 | X = 1) = 0.75
```

### Likert mode

Likert mode presents a seven-point verbal scale based on the
[Sherman Kent](https://en.wikipedia.org/wiki/Words_of_estimative_probability)
probability words used in intelligence analysis. Experts choose the option
that best matches their belief, and the package maps it to a representative
probability.

| Key | Label | Probability |
|-----|---------------------------------------|-------------|
| 1 | Almost certainly not | ~5% |
| 2 | Very unlikely | ~15% |
| 3 | Unlikely | ~30% |
| 4 | About as likely as not | ~50% |
| 5 | Likely | ~70% |
| 6 | Very likely | ~85% |
| 7 | Almost certain | ~95% |

A decimal value can always be typed instead of a key, giving experts the
flexibility to be more precise when they want to.

```{r likert-mode}
result_likert <- elicit_dag_priors(
  dag_xy,
  mode   = "likert",
  target = "Y",
  .responses = list(
    X = list("3"),       # "Unlikely" -> 0.30
    Y = list("2", "6")   # X=0: "Very unlikely" -> 0.15 | X=1: "Very likely" -> 0.85
  )
)
```

---

## Common DAG structures

### Chain: X → Y → Z

In a chain, $X$ influences $Z$ only *through* $Y$. Conditioning on $Y$
makes $X$ and $Z$ independent (the Markov condition).

```{r chain}
dag_chain <- dagitty("dag { X -> Y -> Z }")

result_chain <- elicit_dag_priors(
  dag_chain,
  target     = "Z",
  verbose    = FALSE,   # suppress output — we'll print manually
  .responses = list(
    X = list(0.40),          # P(X = 1) = 0.40
    Y = list(0.15, 0.80),    # P(Y=1|X=0) = 0.15, P(Y=1|X=1) = 0.80
    Z = list(0.20, 0.90)     # P(Z=1|Y=0) = 0.20, P(Z=1|Y=1) = 0.90
  )
)

print(result_chain)
```

We can verify the Markov condition by checking that $X \perp\!\!\!\perp Z
\mid Y$:

```{r chain-markov}
# P(Z=1 | Y=1)  —  should not change when we also condition on X
p_z_y1       <- get_conditional(result_chain, c(Z = 1L), c(Y = 1L))
p_z_y1_x0    <- get_conditional(result_chain, c(Z = 1L), c(Y = 1L, X = 0L))
p_z_y1_x1    <- get_conditional(result_chain, c(Z = 1L), c(Y = 1L, X = 1L))

cat(sprintf("P(Z=1 | Y=1)         = %.4f\n", p_z_y1))
cat(sprintf("P(Z=1 | Y=1, X=0)    = %.4f\n", p_z_y1_x0))
cat(sprintf("P(Z=1 | Y=1, X=1)    = %.4f\n", p_z_y1_x1))
```

All three values match — $X$ carries no additional information about $Z$
once $Y$ is known.

### Fork: Z → X, Z → Y (common cause)

A fork (or "common cause") structure creates a spurious association
between $X$ and $Y$ through their shared parent $Z$. Conditioning on $Z$
removes that association.

```{r fork}
dag_fork <- dagitty("dag { Z -> X ; Z -> Y }")

result_fork <- elicit_dag_priors(
  dag_fork,
  verbose    = FALSE,
  .responses = list(
    Z = list(0.50),
    X = list(0.10, 0.90),   # X is strongly driven by Z
    Y = list(0.10, 0.90)    # Y is equally strongly driven by Z
  )
)

# Marginal association between X and Y (should be non-zero due to Z)
mi_xy_marginal <- get_mutual_info(result_fork, "X", "Y")
cat(sprintf("I(X ; Y)         = %.4f bits\n", mi_xy_marginal))

# After conditioning on Z=0: should be close to zero
p_y1_x1_z0 <- get_conditional(result_fork, c(Y = 1L), c(X = 1L, Z = 0L))
p_y1_x0_z0 <- get_conditional(result_fork, c(Y = 1L), c(X = 0L, Z = 0L))
cat(sprintf("P(Y=1|X=1,Z=0) = %.4f\n", p_y1_x1_z0))
cat(sprintf("P(Y=1|X=0,Z=0) = %.4f\n", p_y1_x0_z0))
cat("(These should match — X provides no info about Y once Z is known)\n")
```

### Collider: X → Z ← Y

A collider is the reverse: $Z$ has two causes but those causes are
otherwise unrelated. Unlike chains and forks, conditioning **on** the
collider $Z$ *opens* a path between $X$ and $Y$, inducing a spurious
association even though none existed marginally.

```{r collider}
dag_coll <- dagitty("dag { X -> Z ; Y -> Z }")

result_coll <- elicit_dag_priors(
  dag_coll,
  target     = "Z",
  verbose    = FALSE,
  .responses = list(
    X = list(0.50),
    Y = list(0.50),
    # Z = 1 only if both causes are present
    Z = list(0.02, 0.30, 0.30, 0.95)
    # order: P(Z=1|X=0,Y=0), P(Z=1|X=1,Y=0),
    #        P(Z=1|X=0,Y=1), P(Z=1|X=1,Y=1)
  )
)

# Marginal: X and Y are independent
mi_xy  <- get_mutual_info(result_coll, "X", "Y")
cat(sprintf("I(X ; Y) marginal     = %.6f bits (should be ~0)\n", mi_xy))

# Conditional on Z=1: X and Y become associated ("explaining away")
p_x1_y1_z1 <- get_conditional(result_coll, c(X = 1L), c(Y = 1L, Z = 1L))
p_x1_y0_z1 <- get_conditional(result_coll, c(X = 1L), c(Y = 0L, Z = 1L))
cat(sprintf("P(X=1 | Y=1, Z=1)    = %.4f\n", p_x1_y1_z1))
cat(sprintf("P(X=1 | Y=0, Z=1)    = %.4f\n", p_x1_y0_z1))
cat("(X becomes less probable when Y is known to be 1 and Z is known to be 1)\n")
```

---

## Querying the model

### Marginal distributions

`get_marginal()` sums the joint distribution over all other variables.

```{r marginals}
get_marginal(result_chain, "X")
get_marginal(result_chain, "Z")
```

### Conditional probabilities

`get_conditional()` accepts named integer vectors for both the query and
the evidence. Conditions on any subset of nodes.

```{r conditionals}
# P(Z = 1 | X = 1)  — propagating belief forward through the chain
get_conditional(result_chain, query = c(Z = 1L), evidence = c(X = 1L))

# P(X = 1 | Z = 1)  — updating belief about a cause given an effect
get_conditional(result_chain, query = c(X = 1L), evidence = c(Z = 1L))
```

### Mutual information

`get_mutual_info()` measures how much knowing one variable reduces
uncertainty about another. It is always non-negative and equals zero for
independent variables.

```{r mutual-info}
# Chain: information degrades along the chain
mi_xz <- get_mutual_info(result_chain, "X", "Z")
mi_xy <- get_mutual_info(result_chain, "X", "Y")
mi_yz <- get_mutual_info(result_chain, "Y", "Z")

cat(sprintf("I(X ; Y) = %.4f bits\n", mi_xy))
cat(sprintf("I(Y ; Z) = %.4f bits\n", mi_yz))
cat(sprintf("I(X ; Z) = %.4f bits   (weakest — one hop further)\n", mi_xz))
```

The data processing inequality guarantees $I(X; Z) \leq \min\{I(X; Y),
I(Y; Z)\}$ in a chain, which we can verify:

```{r dpi}
stopifnot(mi_xz <= min(mi_xy, mi_yz) + 1e-9)
cat("Data processing inequality holds.\n")
```

---

## Entropy interpretation

Shannon entropy $H = -\sum_i p_i \log_2 p_i$ (in bits) measures
uncertainty in the joint distribution.

- **Maximum entropy** ($H = n$ bits for $n$ binary nodes) arises when all
  $2^n$ joint states are equally probable — maximum ignorance.
- **Minimum entropy** ($H = 0$) means only one state has nonzero
  probability — complete certainty.
- The **entropy ratio** $H / H_{\max}$ (printed as "% of maximum") is a
  convenient summary: values near 100% indicate that the elicited
  probabilities leave a great deal of uncertainty, while low values reflect
  strong determinism in the model.

```{r entropy-example}
# A very deterministic chain: X -> Y -> Z with near-certain links
result_det <- elicit_dag_priors(
  dag_chain, verbose = FALSE,
  .responses = list(
    X = list(0.5),
    Y = list(0.02, 0.98),
    Z = list(0.02, 0.98)
  )
)

# A very uncertain chain: links are all near 50 / 50
result_unc <- elicit_dag_priors(
  dag_chain, verbose = FALSE,
  .responses = list(
    X = list(0.5),
    Y = list(0.45, 0.55),
    Z = list(0.45, 0.55)
  )
)

cat(sprintf("Deterministic model  H = %.4f bits  (%.1f%% of max)\n",
            result_det$entropy,
            100 * result_det$entropy / log2(8)))

cat(sprintf("Uncertain model      H = %.4f bits  (%.1f%% of max)\n",
            result_unc$entropy,
            100 * result_unc$entropy / log2(8)))
```

The per-node marginal entropies stored in `result$node_entropies` are
useful for identifying which variables contribute most to overall
uncertainty.

```{r node-entropies}
result_chain$node_entropies
```

---

## A worked applied example

Suppose you are modelling whether a randomised controlled trial ($T$,
treatment) leads to a positive outcome ($O$) through a biological pathway
($M$, mediator), with unmeasured confounding captured by a background
factor ($U$).

```{r applied-dag}
dag_rct <- dagitty("dag {
  T -> M -> O
  U -> M
  U -> O
}")
```

```{r applied-elicit}
result_rct <- elicit_dag_priors(
  dag_rct,
  target  = "O",
  verbose = FALSE,
  .responses = list(
    T = list(0.50),          # balanced randomisation
    U = list(0.30),          # 30% have the background risk factor
    M = list(                # P(M=1 | T, U): 4 combinations
      0.05,   # T=0, U=0
      0.20,   # T=1, U=0
      0.30,   # T=0, U=1
      0.60    # T=1, U=1
    ),
    O = list(                # P(O=1 | M, U): 4 combinations
      0.10,   # M=0, U=0
      0.70,   # M=1, U=0
      0.25,   # M=0, U=1
      0.85    # M=1, U=1
    )
  )
)
```

```{r applied-queries}
# Intent-to-treat effect: P(O=1|T=1) vs P(O=1|T=0)
p_o_t1 <- get_conditional(result_rct, c(O = 1L), c(T = 1L))
p_o_t0 <- get_conditional(result_rct, c(O = 1L), c(T = 0L))
cat(sprintf("P(O=1 | T=1) = %.4f\n", p_o_t1))
cat(sprintf("P(O=1 | T=0) = %.4f\n", p_o_t0))
cat(sprintf("Risk difference = %.4f\n", p_o_t1 - p_o_t0))

# How much of the variation in O is explained by M?
cat(sprintf("\nI(M ; O) = %.4f bits\n", get_mutual_info(result_rct, "M", "O")))
cat(sprintf("I(T ; O) = %.4f bits\n", get_mutual_info(result_rct, "T", "O")))
cat(sprintf("I(U ; O) = %.4f bits\n", get_mutual_info(result_rct, "U", "O")))

# Joint entropy summary
cat(sprintf("\nJoint entropy: %.4f / %.4f bits (%.1f%%)\n",
            result_rct$entropy,
            log2(2^4),
            100 * result_rct$entropy / log2(2^4)))
```

---

## Converting to a CausalQueries model

`to_causalqueries()` lifts an `elicit_dag_result` into a
[CausalQueries](https://integrated-inferences.github.io/CausalQueries/)
`causal_model`. This unlocks **potential-outcomes queries** — statements
like "what is the probability that $T$ was the actual cause of $O$ for a
unit that received treatment and recovered?" — that are not expressible
with the joint distribution alone.

```{r cq-load, message=FALSE}
library(CausalQueries)
```

### How the conversion works

A CPT specifies $P(\text{node}=1 \mid \text{parents}=\mathbf{c})$ for
each observed parent combination $\mathbf{c}$. CausalQueries instead
parameterises the model by **nodal types** — complete response functions
$\theta : \{0,1\}^k \to \{0,1\}$ that specify a node's value for *every*
possible parent combination simultaneously. For a node with $k$ parents
there are $2^{2^k}$ such functions.

`to_causalqueries()` converts using the **independence of potential
outcomes** assumption: the responses $Y(\mathbf{c})$ for different parent
combinations are treated as independent Bernoulli draws. This gives

$$P(\theta) = \prod_{\mathbf{c}\,\in\,\{0,1\}^k}
  P\!\left(Y = \theta(\mathbf{c}) \mid \text{parents} = \mathbf{c}\right).$$

For a single binary parent this produces four nodal types:

| Type | Meaning | Probability |
|------|---------|-------------|
| `00` | $Y=0$ regardless of $X$ ("never") | $(1-p_0)(1-p_1)$ |
| `10` | $Y=1$ when $X=0$, $Y=0$ when $X=1$ (contrarian) | $p_0(1-p_1)$ |
| `01` | $Y=0$ when $X=0$, $Y=1$ when $X=1$ (complier) | $(1-p_0)p_1$ |
| `11` | $Y=1$ regardless of $X$ ("always") | $p_0 p_1$ |

where $p_0 = P(Y=1\mid X=0)$ and $p_1 = P(Y=1\mid X=1)$. Crucially, the
CPT marginals are always exactly recovered:
$P(Y=1\mid X=0) = P(\theta_{10}) + P(\theta_{11}) = p_0$.

### Inspecting the parameter table

```{r cq-xy}
cq_xy <- to_causalqueries(result_xy)
cq_xy$parameters_df[, c("param_names", "nodal_type", "param_value")]
```

The `param_value` column now holds the nodal-type probabilities. For $Y$:
complier (`Y.01`) is by far the most probable type because the elicited
probabilities ($p_0 = 0.10$, $p_1 = 0.75$) strongly favour a positive
effect.

```{r cq-xy-check}
# Recover P(Y=1|X=1) = P(Y.01) + P(Y.11)
with(
  subset(cq_xy$parameters_df, node == "Y"),
  sum(param_value[nodal_type %in% c("01", "11")])
)
```

### Causal queries on the RCT model

Convert the applied RCT result from the previous section:

```{r cq-rct, warning=FALSE}
cq_rct <- to_causalqueries(result_rct)
```

**Average treatment effect** — probability that treatment causes a
positive outcome:

```{r cq-ate, warning=FALSE}
query_model(
  cq_rct,
  query  = "O[T=1] > O[T=0]",
  using  = "parameters"
)
```

**Probability of harm** — probability that treatment *prevents* recovery:

```{r cq-harm, warning=FALSE}
query_model(
  cq_rct,
  query  = "O[T=1] < O[T=0]",
  using  = "parameters"
)
```

**Natural indirect effect through the mediator** — probability that
the treatment's benefit operates *via* $M$:

```{r cq-nie, warning=FALSE}
# Compare O under T=1 vs the counterfactual where M is held at M[T=0]
query_model(
  cq_rct,
  query  = "O[T=1] > O[T=1, M=M[T=0]]",
  using  = "parameters"
)
```

The indirect effect query isolates the component of the treatment effect
that runs through the biological pathway $M$: it compares what the outcome
would be under treatment to what it would have been under treatment *had
the mediator not changed*. A positive value indicates that the pathway
$T \to M \to O$ carries part of the benefit.

### Updating with observed data

Once you have collected data, CausalQueries can update the nodal-type
distribution using `update_model()`, treating the elicited probabilities
as a prior:

```{r cq-update, eval=FALSE}
# Simulate 200 observations from the model (for illustration)
set.seed(42)
obs_data <- simulate_data(cq_rct, n = 200)

# Update: elicited priors + observed data -> posterior
cq_updated <- update_model(cq_rct, obs_data)

# Re-run the ATE query on the posterior
query_model(cq_updated, query = "O[T=1] > O[T=0]", using = "posteriors")
```

This Bayesian updating workflow makes `to_causalqueries()` particularly
useful for **prior elicitation**: you encode domain knowledge before data
collection, then update once evidence arrives.

---

## Tips for effective elicitation

**Decompose before you elicit.** A node with $k$ parents requires
$2^k$ probability assessments. For $k \geq 4$ consider whether the DAG
can be restructured so that no node has more than three parents.

**Use Likert mode for non-statisticians.** When working with domain
experts unfamiliar with probability notation, the verbal scale reduces
cognitive load and tends to produce more internally consistent responses.
Experts can still override with a decimal when they want to be precise.

**Use `.responses` for reproducibility.** Embed your final elicited values
in a script using the `.responses` argument so analyses can be re-run
without re-prompting.

**Entropy as a calibration check.** An entropy ratio above ~90% often
indicates that the elicited probabilities are all close to 0.5, which may
mean the expert genuinely has little information — or that the questions
were too difficult. Review the CPTs and consider simplifying the model.

**Sensitivity analysis.** Run `elicit_dag_priors()` multiple times with
slightly perturbed probabilities to see how sensitive your conclusions are
to the specific values provided. The `.responses` argument makes this
straightforward to script.
