---
title: "The Multilinear Probability Transform"
subtitle: "Encoding Conditional Probability Tables as Polynomial Forms"
author: "Robert Kubinec"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    highlight-style: github
    html-math-method: mathjax
vignette: >
  %\VignetteIndexEntry{The Multilinear Probability Transform}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(elicitcausal)
library(dagitty)
```

## Overview

`to_theorytools()` converts an elicited conditional probability table (CPT)
into a *form* expression for the
[theorytools](https://cjvanlissa.github.io/theorytools/) augmented DAG format.
The conversion relies on a mathematical object called the **multilinear
probability polynomial**, which is the unique polynomial that
(a) exactly matches every entry of the CPT when the parent variables take
binary values, and (b) is linear in each parent variable separately.

This vignette derives the transform step by step, works through one- and
two-parent examples with concrete numbers, and shows how the coefficients
relate to the structural equation that theorytools uses for simulation and
estimation.

---

## 1  Setup and notation

Consider a binary child node $Y \in \{0, 1\}$ with $k$ binary parent nodes
$X_1, X_2, \ldots, X_k \in \{0,1\}$.

A **conditional probability table (CPT)** for $Y$ is a function that assigns
a probability to every combination of parent values:

$$
p_{\mathbf{c}} \;=\; P(Y = 1 \mid X_1 = c_1, \ldots, X_k = c_k),
\qquad \mathbf{c} = (c_1, \ldots, c_k) \in \{0,1\}^k.
$$

There are $2^k$ such combinations, so the CPT contains $2^k$ numbers, each
in $[0, 1]$.

---

## 2  The multilinear polynomial

We seek a function $f : \mathbb{R}^k \to \mathbb{R}$ such that

$$
f(c_1, \ldots, c_k) = p_{\mathbf{c}}
\qquad \text{for every } \mathbf{c} \in \{0,1\}^k.
$$

Among all functions satisfying this interpolation condition, we choose the
one that is **multilinear** — linear (i.e., at most degree-1) in each
variable $X_i$ separately when all other variables are held fixed.

**Claim.** There is a unique multilinear polynomial satisfying the
interpolation condition, and it takes the form

$$
\boxed{f(x_1, \ldots, x_k) \;=\;
  \sum_{S \subseteq \{1,\ldots,k\}} \alpha_S \prod_{i \in S} x_i,}
$$

where the sum runs over all $2^k$ subsets $S$ of the parent index set
(including $S = \emptyset$, whose monomial is the empty product, equal to
$1$).

*Proof of uniqueness.* The $2^k$ monomials $\prod_{i \in S} x_i$, indexed by
subsets $S \subseteq \{1,\ldots,k\}$, are linearly independent on
$\{0,1\}^k$ (the evaluation map at any $\mathbf{c}$ picks out a unique
lower-triangular system — see Section 3).  Hence the expansion is unique.

*Proof of multilinearity.* Each variable $x_j$ appears at most once in every
monomial (since $S$ is a set, not a multiset), so $f$ is linear in each
$x_j$ separately. $\square$

---

## 3  Computing the coefficients: the Möbius transform

### 3.1  Design matrix formulation

Order the $2^k$ binary combinations $\mathbf{c}^{(1)}, \ldots, \mathbf{c}^{(2^k)}$
using `expand.grid` with the **first parent cycling fastest**:

$$
\mathbf{c}^{(1)} = (0,0,\ldots,0), \quad
\mathbf{c}^{(2)} = (1,0,\ldots,0), \quad
\mathbf{c}^{(3)} = (0,1,\ldots,0), \quad \ldots
$$

Stack the CPT probabilities into a vector:

$$
\mathbf{p} = \bigl(p_{\mathbf{c}^{(1)}}, \ldots, p_{\mathbf{c}^{(2^k)}}\bigr)^\top \in [0,1]^{2^k}.
$$

Order the $2^k$ subsets $S$ correspondingly (also by `model.matrix` with
the formula `~ X1 * X2 * ... * Xk`) and stack the coefficients:

$$
\boldsymbol{\alpha} = (\alpha_\emptyset,\, \alpha_{\{1\}},\, \alpha_{\{2\}},\,
  \alpha_{\{1,2\}},\, \ldots)^\top \in \mathbb{R}^{2^k}.
$$

The interpolation condition $f(\mathbf{c}^{(i)}) = p_{\mathbf{c}^{(i)}}$
for all $i$ becomes the linear system

$$
\mathbf{X}\,\boldsymbol{\alpha} = \mathbf{p},
$$

where $\mathbf{X} \in \{0,1\}^{2^k \times 2^k}$ is the model matrix with

$$
X_{i,S} = \prod_{j \in S} c^{(i)}_j.
$$

### 3.2  The matrix $\mathbf{X}$ is always invertible

For binary predictors, $\mathbf{X}$ is a lower-triangular matrix with 1s on
the diagonal (the row corresponding to $\mathbf{c} = \mathbf{0}$ has only
the intercept column nonzero; the rows with more 1s progressively fill in
more columns).  Its determinant is therefore $1 \neq 0$, so

$$
\boldsymbol{\alpha} = \mathbf{X}^{-1}\,\mathbf{p}
$$

always has a unique solution.  The inverse $\mathbf{X}^{-1}$ is the
**Möbius inversion matrix** for the Boolean lattice.

### 3.3  Closed-form expression

The Möbius inversion gives a closed-form formula for each coefficient:

$$
\alpha_S \;=\; \sum_{T \subseteq S} (-1)^{|S| - |T|}\, p_T,
$$

where $p_T$ denotes the CPT entry with parents set to 1 exactly on the
indices in $T$ (and 0 elsewhere).  The alternating-sign inclusion–exclusion
structure is the Boolean analog of finite-difference operators.

In `elicitcausal`, the coefficients are computed via `stats::model.matrix`
and `base::solve`:

```{r coef-computation, eval=FALSE}
X_design <- stats::model.matrix(
  stats::as.formula(paste0("~ ", paste(parent_names, collapse = " * "))),
  data = expand.grid(lapply(parent_names, function(.) c(0L, 1L)),
                     KEEP.OUT.ATTRS = FALSE)
)
alpha <- solve(X_design, cpt_probs)
```

---

## 4  Single-parent example ($k = 1$)

Let $Y$ have one parent $X$, with

$$
p_0 = P(Y=1 \mid X=0), \qquad p_1 = P(Y=1 \mid X=1).
$$

The design matrix and system are

$$
\underbrace{\begin{pmatrix} 1 & 0 \\ 1 & 1 \end{pmatrix}}_{\mathbf{X}}
\begin{pmatrix} \alpha_\emptyset \\ \alpha_{\{1\}} \end{pmatrix}
=
\begin{pmatrix} p_0 \\ p_1 \end{pmatrix},
$$

giving

$$
\alpha_\emptyset = p_0, \qquad \alpha_{\{1\}} = p_1 - p_0.
$$

The multilinear polynomial is therefore

$$
f(x) = p_0 + (p_1 - p_0)\,x.
$$

This is a simple linear interpolation between the two CPT entries.  The
theorytools `form` on the edge $X \to Y$ is:

$$
\texttt{form} = \underbrace{p_0}_{\text{intercept}} + \underbrace{(p_1 - p_0)}_{\text{slope}} \cdot X.
$$

**Concrete example:** $p_0 = 0.2$, $p_1 = 0.8$:

$$
f(x) = 0.2 + 0.6\,x.
$$

Verify: $f(0) = 0.2\,\checkmark$, $f(1) = 0.8\,\checkmark$.

```{r single-parent-ex}
result_xy <- elicit_dag_priors(
  dagitty("dag { X -> Y }"),
  verbose    = FALSE,
  .responses = list(X = list(0.5), Y = list(0.2, 0.8))
)
addag_xy <- to_theorytools(result_xy)
cat(as.character(addag_xy))
```

---

## 5  Two-parent example ($k = 2$)

Let $Y$ have parents $X_1$ and $X_2$, with CPT entries
$p_{00}, p_{10}, p_{01}, p_{11}$ (first parent cycles fastest):

$$
\mathbf{X} =
\begin{pmatrix}
1 & 0 & 0 & 0 \\
1 & 1 & 0 & 0 \\
1 & 0 & 1 & 0 \\
1 & 1 & 1 & 1
\end{pmatrix},
\qquad
\mathbf{p} =
\begin{pmatrix} p_{00} \\ p_{10} \\ p_{01} \\ p_{11} \end{pmatrix}.
$$

Solving $\mathbf{X}\boldsymbol{\alpha} = \mathbf{p}$:

$$
\begin{aligned}
\alpha_\emptyset   &= p_{00}, \\
\alpha_{\{1\}}     &= p_{10} - p_{00}, \\
\alpha_{\{2\}}     &= p_{01} - p_{00}, \\
\alpha_{\{1,2\}}   &= p_{11} - p_{10} - p_{01} + p_{00}.
\end{aligned}
$$

The multilinear polynomial is

$$
f(x_1, x_2)
  = p_{00}
  + (p_{10} - p_{00})\,x_1
  + (p_{01} - p_{00})\,x_2
  + (p_{11} - p_{10} - p_{01} + p_{00})\,x_1 x_2.
$$

The coefficient of the interaction term $x_1 x_2$ is the **second-order
finite difference** of the CPT — it is zero whenever the effect of $X_1$ on
$Y$ is the same regardless of $X_2$'s value (i.e., when the two parents act
independently on $Y$).

**Concrete example:** collider $X \to Z \leftarrow Y$ with
$p_{00} = 0.1$, $p_{10} = 0.4$, $p_{01} = 0.3$, $p_{11} = 0.9$:

$$
\begin{aligned}
\alpha_\emptyset  &= 0.1, \\
\alpha_{\{X\}}   &= 0.3, \\
\alpha_{\{Y\}}   &= 0.2, \\
\alpha_{\{X,Y\}} &= 0.9 - 0.4 - 0.3 + 0.1 = 0.3.
\end{aligned}
$$

$$
f(x, y) = 0.1 + 0.3\,x + 0.2\,y + 0.3\,xy.
$$

```{r two-parent-ex}
result_xzy <- elicit_dag_priors(
  dagitty("dag { X -> Z; Y -> Z }"),
  verbose    = FALSE,
  .responses = list(X = list(0.5), Y = list(0.5),
                    Z = list(0.1, 0.4, 0.3, 0.9))
)
addag_xzy <- to_theorytools(result_xzy)
cat(as.character(addag_xzy))
```

We can verify all four CPT entries by evaluating the combined edge forms:

```{r two-parent-verify}
# Substitute : -> * so R evaluates interaction as multiplication
form_eval <- function(s, ...) eval(parse(text = gsub(":", "*", s)), list(...))

dag_str <- as.character(addag_xzy)

get_form <- function(pattern, text) {
  m <- regmatches(text, regexpr(pattern, text))
  sub(pattern, "\\1", m)
}

form_X <- get_form('X -> Z \\[form="([^"]+)"\\]', dag_str)
form_Y <- get_form('Y -> Z \\[form="([^"]+)"\\]', dag_str)

combos <- expand.grid(X = 0:1, Y = 0:1)
combos$prob <- mapply(
  function(x, y) form_eval(form_X, X = x, Y = y) + form_eval(form_Y, X = x, Y = y),
  combos$X, combos$Y
)
combos
```

---

## 6  Edge attribution rule

The theorytools format attaches a `form` to each **directed edge**, not to
the child node as a whole.  When a child has multiple parents, the
$2^k$ monomials of the polynomial must be distributed across the $k$
incoming edges.

`to_theorytools()` uses the following rule:

> **Each monomial $\alpha_S \prod_{i \in S} x_i$ is attributed to the edge
> of the parent with the smallest index in $S$.  The constant term
> $\alpha_\emptyset$ is attributed to the first parent's edge.**

Concretely, for $k = 2$ parents $X_1, X_2$:

| Monomial | Attributed to edge |
|---|---|
| $\alpha_\emptyset$ (constant) | $X_1 \to Y$ |
| $\alpha_{\{1\}} x_1$ | $X_1 \to Y$ |
| $\alpha_{\{2\}} x_2$ | $X_2 \to Y$ |
| $\alpha_{\{1,2\}} x_1 x_2$ | $X_1 \to Y$ (min index is 1) |

The two edge forms therefore are:

$$
\begin{aligned}
\text{form}(X_1 \to Y) &= \alpha_\emptyset + \alpha_{\{1\}}\,X_1 + \alpha_{\{1,2\}}\,X_1{:}X_2, \\
\text{form}(X_2 \to Y) &= \alpha_{\{2\}}\,X_2.
\end{aligned}
$$

Summing them recovers $f(x_1, x_2)$ exactly.  The rule ensures that every
monomial appears on exactly one edge: the edge of the parent most responsible
for "introducing" the term.

---

## 7  Relationship to logistic regression

The multilinear polynomial lives in **probability space** (the output of $f$
is directly $P(Y=1|\ldots)$), not in log-odds space.  The standard logistic
regression model uses a logit link:

$$
\log\frac{P(Y=1|\mathbf{x})}{1-P(Y=1|\mathbf{x})} = \beta_0 + \sum_i \beta_i x_i + \sum_{i<j} \beta_{ij} x_i x_j + \cdots
$$

The logistic-regression coefficients $\boldsymbol{\beta}$ are **not** the
same as the multilinear coefficients $\boldsymbol{\alpha}$, but both
parameterize the same CPT exactly (since for binary predictors, a saturated
logistic model also has $2^k$ free parameters).

The multilinear formulation has two practical advantages over the logistic
one for this application:

1. **Direct simulation.** Because $f(\mathbf{x}) \in [0,1]$ for all
   $\mathbf{x} \in [0,1]^k$, theorytools' `simulate_data()` can draw
   $$Y \sim \text{Bernoulli}(f(x_1, \ldots, x_k))$$
   directly without a link-function transformation.

2. **Closed form.** The coefficients $\boldsymbol{\alpha}$ are obtained by a
   single matrix solve with an integer matrix (no iterative optimization
   required), and the solution is always exact.

For **statistical estimation** from observed data, a logistic regression
(binomial family, logit link) is generally preferred for numerical stability:

```r
glm(Y ~ X1 * X2, data = df, family = binomial)
```

The identity-link binomial model matches the multilinear polynomial exactly:

```r
glm(Y ~ X1 * X2, data = df, family = binomial(link = "identity"))
```

though the identity-link solver can be numerically unstable when predicted
probabilities approach 0 or 1.

---

## 8  General $k$-parent formula

For completeness, the general coefficient formula via Möbius inversion is:

$$
\alpha_S = \sum_{T \subseteq S} (-1)^{|S|-|T|}\, p_T,
$$

where $p_T = P(Y=1 \mid X_i = \mathbf{1}[i \in T])$ is the CPT entry with
parents in $T$ set to 1 and all others set to 0.

**Special cases:**

- $|S| = 0$: $\alpha_\emptyset = p_\emptyset = p_{00\cdots0}$ (baseline probability).
- $|S| = 1$, $S = \{j\}$: $\alpha_{\{j\}} = p_{\{j\}} - p_\emptyset$ (marginal increment from $X_j$).
- $|S| = 2$, $S = \{i,j\}$: $\alpha_{\{i,j\}} = p_{\{i,j\}} - p_{\{i\}} - p_{\{j\}} + p_\emptyset$ (interaction).

The interaction coefficient $\alpha_{\{i,j\}}$ is zero when the causal effects
of $X_i$ and $X_j$ on $Y$ are **additive on the probability scale** — i.e.,
when $P(Y=1|X_i=1,X_j=1) = P(Y=1|X_i=1,X_j=0) + P(Y=1|X_i=0,X_j=1) - P(Y=1|X_i=0,X_j=0)$.

---

## 9  Summary

| Quantity | Symbol | Meaning |
|---|---|---|
| CPT entry | $p_\mathbf{c}$ | $P(Y=1 \mid \text{parents} = \mathbf{c})$, elicited from user |
| Multilinear polynomial | $f(x_1,\ldots,x_k)$ | Unique interpolant; equals $p_\mathbf{c}$ at every binary input |
| Möbius coefficient | $\alpha_S$ | Coefficient of monomial $\prod_{i\in S} x_i$ |
| Design matrix | $\mathbf{X}$ | `model.matrix(~X1*...*Xk)` at binary combinations |
| Linear system | $\mathbf{X}\boldsymbol{\alpha} = \mathbf{p}$ | Solved exactly; $\mathbf{X}$ is always invertible |
| theorytools `distribution` | `rbinom(size=1)` | Signals binary Bernoulli response family |
| theorytools `form` | $f$ split across edges | Each edge carries the monomials of its lowest-indexed parent |
